Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/example.tex

@@

In this paper, we develop a splitting algorithm
(\ref{sec:splitting,sec:implementation}) which, given a program $e$ in \lang,
produces an equivalent pair of programs which correspond precisely to the two
stages of computation in $e$.
(Splitting is always possible because the staging annotations in $e$ are
consistent, because $e$ is well-typed in \lang.) In the case of
\texttt{qsStaged}, our splitting algorithm produces the algorithm described
above.

%\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.

@@

>>>>

It is well-known that for any given input list with $n$ elements, the
quick-select algorithm requires $\Theta(n)$ expected time (for any
second argument) if we select keys for the partition function
uniformly randomly or identically if we permute the input. Under this
assumption, our example program requires expected $\Theta(nm)$ time.

Suppose now that $m$ is large (close to $n$ for example).  In this
case, our program is essentially a quadratic-time algorithm, which can
be very slow even for modest values of $m$ and $n$. In this case, to
improve the run-time it appears that we have no choice but rewrite our
program to use a different algorithm.  One way to do this would be to
pre-sort the input list $l$ and then lookup the element with the
desired rank (as specified by the second argument).  While this
approach might sound sensible, attaining the desired performance
requires answering one more important question: what should the data
structure for storing the sorted input be?  Unfortunately using a list
does not work, because each look up in a list requires linear time on
average. One option is to use arrays but in this case, we would need
to implement a binary search to perform the lookup.  Another option is
to use a balanced binary search tree data structure, which supports
lookups in logarithmic time.  Using either one of these techniques, we
can reduce the run-time to $O(n\log{n} + m\log{n})$, which is
significantly better that $O(n^2)$.

In summary, to achieve the desired improvement, we wrote a two-pass
program, where the first pass sorts the input and the second pass
performs a binary search either on an array or a balanced tree. As is
probably apparent to the reader, the resulting program is both
algorithmically and structurally more complex: it requires conversions
between different data structures (list to array or binary search
tree) and requires implementing additional algorithms such as sorting
and binary search, also possibly binary search trees.  As suggested by
this relatively small example and experienced by the graphics
community, the design, implementation, and maintenance of such
multi-pass programs can be difficult due to the increased complexity.

Instead of relying an writing multipass programs manually, in this
paper, we present techniques for automatically splitting staged
programs into multi-pass programs.  The key innovation behind our
approach is a {\em splitting algorithm} that splits a multistage
program into separate passes where each pass is a conventional program
expressed in a conventional functional language.  Our splitting
algorithm makes no restrictions on the input program: essentially any
functional program, including higher order programs, can be split.
Furthermore, the splitting algorithm is able to perform highly
non-trivial transformations on the program that can improve efficiency
significantly.

When applied to the quick-select example as described, our splitting
algorithm yields a two-pass program that uses the binary-search-tree
based implementation described above.  Specifically, in the first
pass, the program takes the input list and constructs a
probabilistically balanced binary search tree, which is isomorphic to
a treap data structure~\cite{treaps}.  In the second pass, the program
performs a binary search tree lookup, by walking the tree to find the
element with the desired rank.  The resulting program takes the
$O(n\log{n} + m\log{n})$.

We note that the splitting algorithm operates by composing local
transformations on the subterms of the input program.  In particular,
the algorithm has no special knowledge of the quick-select algorithm,
binary search trees, or how to perform lookups on binary search trees,
but it is able to derive all of these from the input program.  But
how?

The technical detail of the algorithm are relatively involved but in
the hope that it aids understanding we present a brief informal
overview of how the algorithm operates.  This exposition is
necessarily imprecise but the algorithm is made precise in
\ref{sec:splitting} and implemented
(\ref{sec:implementation}{sec:examples}).

At a high level, the splitting algorithm operates on our example by
using the code for quickselect and the knowledge that the list (the
first argument) is a first-stage argumint and the rank (second
argument) is a second stage argumunt, the algorithm then derives two
functions: \texttt{qsStage1} and \texttt{qsStage2} as follows.

\paragraph{Pass-1 code.}
The splitting algorithm ``imitates'' an execution of quick-select with
just the first-stage input by evaluating as much code as possible with
the available information.  Specifically the algorithm generates a
function \texttt{qsStage1} to perform all the recursive calls of
\texttt{qs} and to evaluate all instances of the partition function,
which depend only on the input list.  The function \texttt{qsStage1}
produces a trace of the execution of \texttt{qs} on its first input by
generating a tree that collects the results from all recursive calls
along with a tag that indicates the control branches taken.  The trace
ends up being represented as a tree because \texttt{qs} has a binary
control structure casing on the list. In fact, function
\texttt{qsStage1} constructs a binary search tree, keyed by the
``pivot'' used by the partition, because the ``pivot'' is needed in
the second stage to construct the result.  We call the result data
structure of \texttt{qsStace1} the {\em boundary}, as it passes
information from the first pass to the second.

\paragraph{Pass-2 code}
As the splitting algorithm ``scans'' \texttt{qs} for computations that
can be performed in the first stage and collects them into
\texttt{qsStage1}, it also collects computations that must be left to
the second stage in a separate function \texttt{qsStage2}.  This
function, which is executed in the second pass, takes as argument the
boundary and the second-stage argument (the rank) and performs a
lookup in the boundary data structure.  But how does \texttt{qsStage2}
knows to traverse the tree? It does not. Since the splitting algorithm
recorded the control flow of the first stage execution in the boundary
data structure, \texttt{qsStage2} simply follows this control
structure and performs at each point the parts of the computation from
\texttt{qs} that can now be performed in the presence of the
second-stage argument (the rank).

\begin{comment}

@@

In this paper, we develop a splitting algorithm
(\ref{sec:splitting,sec:implementation}) which, given a program $e$ in \lang,
produces an equivalent pair of programs which correspond precisely to the two
stages of computation in $e$.
(Splitting is always possible because the staging annotations in $e$ are
consistent, because $e$ is well-typed in \lang.) In the case of
\texttt{qsStaged}, our splitting algorithm produces the algorithm described
above.
\end{comment}
%\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.

@@

\subsection{Quickselect in more detail}

Carlo writes this.
